Linking watershed nutrient loading to estuary water quality with Generalized Additive Models

### 2023-08-11 Rebuttal_Letter/Rebuttal_Letter.Rmd

---
author: Michael Schramm
return-address: 
  - Texas Water Resources Institute
  - 1001 Holleman Drive E.
  - College Station, Texas 77840
return-email: michael.schramm@ag.tamu.edu
address:
  - Dr. James Latimer
  - Academic Editor
  - PeerJ Life & Environment

@@

subject: Revised Manuscript
opening: Dear Dr. Latimer,
closing: Best regards,
signature: Michael Schramm

lco: UScommercial9
papersize: letter

inline-links: true

komaoption:
  - parskip=full

@@

Thank you for the prompt and thorough review of my manuscript, "Linking watershed nutrient loading to estuary water quality with semiparametric models." I really appreciate the substantive and helpful comments from each of the reviewers. The final draft and
draft with differences marked are attached. Please note that the latexdiff tool
did not highlight the very minor changes in the abstract or references. I also would like to note the updated title based on comments from one reviewer.

Line by line comments and responses are included below (reviewer comments are italicized).
I have incorporated nearly all of the requested changes from the reviewers.

**Reviewer 1**

**Comment:**
*Very well-structured abstract – introduced the problem, the objectives of the research, the results obtained, and the conclusion. No changes are recommended.*
*The introduction was well-structured and easy to follow. No changes are recommended.*

@@

**Response:** Noted, and thank you.

**Comment:**
*Line 114: A link to the USGS NWIS system could be added.*

**Response:** The url has been added (Line 122).

**Comment:**
*A brief description of the sample collection and laboratory analyses described in the QAPP and TCEQ procedures manual, detection and recovery limits, and instrumentation would seem appropriate.*

@@

**Response:** I added the EPA methods and TCEQ's required AWRL's to Tables 1 and 2. I believe this is sufficient documentation.

**Comment:**
*Also, what are the state-recommended guideline values? Include them in Table 2*

@@

**Response:** I've added the associated state water quality criteria or screening levels in both Tables 1 and 2.

**Comment:**
*Line 371 “has” repeated twice*

**Response:** This is fixed.

**Comment:**
*The author adequately answered the research questions and fulfilled all the stated objectives. The results from the GAM model provided estimates of watershed nutrient loadings and an outlook of nutrient loading's influence on water quality.*
*The whole of the paper was well structured, easy to read, and contributes sufficiently to the existing body of knowledge on estuary anthropogenic pollution and water quality monitoring.*
*Through this paper, the author has created an addition to the body of work on using models for the prediction and assessment of pollution in estuaries and aquatic bodies, as a means of recommending and establishing policy actions for pollution mitigation.*

@@

**Response:** Thank you for the review.

**Reviewer 3**

**Comment:**
*Table 1: It would be helpful if you would include the watershed or station name under the USGS station ID in this table.*

@@

**Response:** I've added the associated watersheds to each USGS station in Table 1.

**Comment:**
*Line 164: Not sure if this is a typo, or if I’m missing what log1p(Q) is.*

**Response:** This was shorthand for *log(1+Q)*. I've changed this to *log(1+Q)* to be consistent with equation 1 (line 178).

**Comment:**
*Line 392, "Limitation" section: I expected this to be a discussion of limitations of the approach like matching daily loads to concentrations, the complex environmental factors that could be involved in chlorophyll and DO trends, changing climate, etc. However, since you focus on the (very valid) problem of limited sampling, you may want to rename this sub-section of the Discussion. Maybe call it "Data Needs" or "Suggested Management Focus" or something like that?*

**Response:**
I agree, this subsection title is updated to "Future Management and Research Needs" (line 398).

**Comment:**
*Line 53-54: Since only one of the trends you mention from Bugica is for a nitrogen species, I suggest rephrasing or expanding this sentence to explain why we’d be concerned about all of the parameters that are increasing (or clarify if we are only concerned about the TKN).*

**Response:**
Added a few lines clarifying the need to assess both inorganic and organic species and phosphorus due to the complex roles they have in contributing to estuary health (lines 53-59).

**Comment:**
*Lines 129-131: It would be helpful if you described how deep the water column is at the estuary sample locations and from what depth the samples that you analyzed were taken. A little more explanation of the estuary dynamics would be helpful too -- is this a stratified or well-mixed system, what is the range in salinity, is there a key season of concern for eutrophication?*

**Response:** I have added additional description of the estuary dynamics (lines 108-111). Specifically mixing, turbidity, and salinity. Given the preliminary nature of this study, there is not a season of concern identified yet, so it is not included. Depth measurements are also detailed in lines 139-143.

**Comment:**
*Lines 194-201: Some discussion is needed of how you dealt with temporal matching of flow and loads to concentrations in the estuary. I see in the discussion (line 373-374) that you matched flow and load estimated for the day of each sample. This approach should be explained in the methods, and I think you need some explanation is needed as to why that would be appropriate and/or how it may impact the results. As you discuss below, you might get better relationships (for Table 4, Lines 244-248) if you either lagged the match or matched an average load and flow from some reasonable period before the estuary sample date. If you had the ability to try some models like that, I think it may be a useful addition.*

**Response:** Dr. Marcus Beck made a similar comment in his review. This is now addressed
by incorporating cumulative flows and loads into the estuary models and further discussed below. See revisions in lines 208-214. Some minor changes to the results were made, but the conclusions are largely consistent with the initial submission.

**Comment:**
*Line 296: Would damming of the Navidad River impact the loads through the Lavaca River watershed? I don’t see why if the data was from the USGS-08164000 site.*

**Response:** You are absolutely correct. I mistakenly wrote this section considering the
combined watershed loads, when these estimates were for the Lavaca River watershed only.
I've removed this and incorporated your statement below (lines 301-325). Thank you very much for catching this!

**Comment:**
*Lines 286-321: I think this discussion is too long comparing the TP yield first to the Dunn study, and then to 3 other studies. The TP yield estimate from your study is in the middle of the others, so you can condense this 2 paragraph discussion and simply say that you fall in the middle of other estimates and state briefly that there are reasons such as land use and flow variability that the estimates vary.*

**Response:** I agree, this section has been considerably condensed. See response above.

**Comment:**
*Lines 360-361: It is probably worthwhile to note that the effect of flow you see in chlorophyll is very likely due to more nutrients coming in with high flows. I wouldn’t want a reader to think that since your flow-adjusted nutrient loads are not linked to chlorophyll that this means there is definitely no impact of nutrients on chlorophyll concentrations. Along those lines, I’d suggest in Line 350-251 you specific that “No relationships between flow-adjusted inorganic nitrogen or TP loadings with chlorophyll-a were observed”*

**Response:** This is one place where we had a slight change in the results due to incorporating cumulative flows and loads. This is reflected in revisions in lines 349-358.

**Reviewer: Marcus Beck**

**Comment:**
*This article describes the use of Generalized Additive Models (GAMs) to estimate nutrient loads entering Lavica Bay from the main inflows. The author uses these models to evaluate the anthropogenic contribution of nutrients to the bay through flow-normalization and to evaluate if these loads are related to long-term variation in estuarine water quality. Overall, this is a valuable paper that further validates the use of GAMs for long-term trend analysis and will have particular value for the management of this system. My comments below relate to either clarification or adjustment of the methods that authors should consider. If pursued, I consider these changes appropriate for a major revision.*

@@

**Response:** I'm highly appreciative of these comments, they have substantially improved the manuscript.

**Comment:**
*I have some concerns about the methods starting on line 184 that describe how the load estimates are related to the estuarine data. First, I strongly encourage the authors to develop these models using cumulative or lagged estimates for the flow and load variables. It appears that these are currently modeled as the estimates for the same day the water quality parameter was sampled. I would not expect much of a relationship between water quality and the independent variables at that time scale. I suspect that the relationships the author shows regarding flow/load with the response reflects some level of temporal autocorrelation between the daily estimates and preceding conditions/loads, where the latter may be a truly causative relationship. I see two potential approaches to remedy the issue. One, the author could test different lagged values for flow and load, although this would require a bit of trial and error. A potentially better solution is to use the aggregate flow volume and load between the sample dates as a sum of the volume and mass, respectively, between each observation of the time series for the response variable. I would expect these aggregate measurements to be much more strongly correlated to the response variables than the daily values that are currently used. This limitation is noted on lines 371-374, so I’m curious why the authors did not use this approach.*

@@

*My second concern about the models in eqns. 5-7 also relates to the flow and load variables. It’s stated on line 199-201 that flow-adjusted load values were used, which makes sense, but I don’t understand how this additional analysis differs from the flow-normalization in the previous section (lines 176 – 179). Why not just use the flow-adjusted loads from the previous models? Once this is clarified (or corrected), it might make more sense to change eqn 6 to include smoothers for ddate, yday, and flow-normalized load, and eqn 7 to included smoothers for ddate, yday, and load. These wouldn’t be “nested” models for AIC comparison, but would perhaps provide a better approach to evaluate loads related to natural or non-natural variation. I don’t see the value in including flow as a separate variable since it’s basically a surrogate for total load.*

**Response:** I am combining my response to both these comments since they are related. These are valid concerns and related to a comment made by Reviewer 3 (indicative that I better make a change here).

(1) I agree that it makes more sense to utilize the flow normalized loading estimates generated in the previous section. The methodology and results have been adjusted to correct this (lines 250-220).

(2) I struggled with the concept of incorporating lag effects while developing these models, it didn't occur to me to utilize aggregated loads/flows. I believe this is an appropriate approach and will incorporate it into the models. However, since the estuary water quality data is not measured at equal intervals, the suggested aggregation method is potentially inappropriate. For example, if a quarterly measurement was missed you can end up aggregating 6 months of loads for a given data point instead of 3 months. Walker et al. (2021) found that after Hurricane Harvey (a massive magnitude storm) water quality impacts in Lavaca Bay and adjacent estuaries were restricted to a relatively short period, somewhere between 20 days and a few months. Based on this info, I am incorporating 1-day lagged, 20-day cumulative loads and flows into this set of models instead of the same day measures. The aggregated flows are still seasonally adjusted (lines 206-214).

(3) In regard to the suggestion for changing equations 6 and 7 to utilize loads and flow-normalized loads instead of flow: I recognize there are a number of valid approaches available here. Since aggregated loads and estuary inflow are surrogates, I don't fully understand the value of replacing the seasonally adjust inflow values for aggregated loads. If I understand the models correctly, both the reviewer suggested approach and the approach used in the draft evaluate the relative variation driven by changes in inflow and anthropogenic changes in nonpoint source loads.

In sum, the results have been slightly revised, but overall conclusions from the study remain the same. In particular, responses of chlorophyll-a at one site, and TKN at two of the sites are changed in response to flow and/or nutrient loads. This are updated in the relevant sections in the results and discussion.

**Comment:**
*I also strongly encourage the author to consider replacing “semiparametric models” with “Generalized Additive Models” in the title. It’s a more accurate description of what’s included in the paper.*

@@

**Response:** Title has been updated.

**Comment:**
*Line 18: Maybe replace “non-natural variation” with “anthropogenic sources”.*

**Response:** Suggested changed made (line 18).

**Comment:**
*Line 61: Accidental line break. *

@@

**Response:** Fixed.

**Comment:**
*Line 77: Suggest rephrasing as “…generalized linear models where the response variable is modeled as the sum of…”*

**Response:** Suggested change made (line 80).

**Comment:**
*Tables 1 and 2: It would be useful to indicate the approximate sampling interval for the water quality parameters. It looks like quarterly. *

**Response:** I'm out of space in the tables, but include this in the text (line 142-143).

**Comment:**
*Line 102: Some more information about the Bay would be helpful. What is the average depth? Any information on residence time? Is the water column mixed, stratified, or seasonally-stratified?*

**Response:** I have not identified any information on residence time. I have added more information on average depth, mixing and stratification of the water column (lines 105-120)

**Comment:**
*Lines 128 – 129: The interpretation of the figures, tables, and text would be much easier if the estuarine sites were simply referred to as “upper”, “mid”, and “lower”. I suggest replacing all station names with these terms throughout the manuscript. *

**Response:**
Fixed throughout.

**Comment:**
*Line 129 – 131: Some additional information about the sampling methods for the estuarine sites could be useful. Where in the water column were these samples taken? Is the water column mixed? I’m thinking specifically about dissolved oxygen. The results show that it was not related to load, but you may have different outcomes depending on surface vs bottom DO. *

**Response:**
Added in lines 139-143.

**Comment:**
*Line 138: Typo “tend”.*

**Comment:**
*Figure 1: Remove “contribution” from the legend. *

@@

**Comment:**
*Figure 1: Remove “contribution” from the legend. *

**Response:** Suggested change made.

**Comment:**
*Eqn 2: I think some additional details are needed in the text below this equation. Why was d set at 0.95? Based on the text, this seems awfully close to 1 or the “mean flow”. Also, the equation includes a lag term, t – t_j, but the size of this lag is not stated, i.e., “some historical day of observation”. Was this set explicitly by the author, and if so, is there a justification? *

**Response:** (1) I have updated the equation to (hopefully) be more clear about
the calculation. For each day in the record, the full record of flows is utilized (an admittedly not efficient calculation, but not an issue for a period of record the size of this data set). (2) I set *d* to 0.95 based on results from Robson and Dourdet (2015) and Zhang and Ball (2017), which is now discussed (lines 159-164).

**Comment:**
*Line 151: It’s not clear to me how fa is a unitless term. The equations show these as simple differences of average streamflow, so wouldn’t the units be in log-cfs? *

**Response:** I worded this incorrectly. I should have phrased this "dimensionless." This is fixed (line 165).

**Comment:**
*Line 169: What percentage of the observations for each parameter were non-detects? I realize that setting the values as ½ the limit can introduce bias, but this may not be a big deal if there aren’t many non-detects. There does appear to be a GAM package that handles non-detects with Tobit models, but it is quite old (by software standards). It may be worth exploring. https://cran.r-project.org/web/packages/cenGAM/index.html*

**Response:** 4% of TP samples were below detection limits and 20% of NO~3~ 
samples fell below detection limits (line 182).
Early in this project 
I explored the use of the Tobit I family using the linked cenGAM package and 
unfortunately found the models result in large overestimation of concentrations.
I similarly explored the use of censored Gamma models in brms with similar 
results. Bergbusch et al. (2021) report similar findings. Conversely, Hayes 
et al. (2020) apply a censored GAM using brms without issue on microcystin 
concentration in lakes. brms appears promising for fitting censored water 
quality data, but I was frankly unable to fit reasonable models for this project
which might be due to sample size or my misunderstanding of model application.

**Comment:**
*Line 236: This might be a personal preference, but “unusual periodic pattern” seems like regular periodicity and not a more random periodicity as one would expect with rain events. Consider rephrasing this text. *

**Response:** Rephrased to simply "periodic pattern" (line 255).

**Comment:**
*Figure 2 caption: Consider adding a sentence like “Values closer to 1 for NSE and r and values closer to 0 for PBIAS represent more ideal goodness-of-fit assessments.”*

@@

**Response:** Incorporated the suggested sentence.

**Comment:**
*Figure 3 caption: Consider adding Lavaca and Navidad with the station name, as in the caption for figure 4. *

**Response:** Added Lavaca and Navidad to the captions.

**Comment:**
*Figure 5 caption: Consider including a reference to Table 3, as it wasn’t immediately clear to me that these models didn’t include flow or load. *

@@

**Response:** Added a reference to Table 3.

**Comment:**
*Table 4: As for my previous comment, I’d replace the site names with “upper”, “mid”, and “lower” and arrange them accordingly. The current arrangement is “mid”, “lower”, and “upper”. Also, it looks like several of the statistical summaries are identical, e.g. the inflow and inflow+load model for TP at TCEQ-13384. Are these simply rounding issues? And if so, are these models really all that different in terms of explained variance? *

@@

**Respones:** I've replaced the site names and reordered the table based on location.
You are correct, the statistical summaries for several models are identical indicating that the more complicated models are not explaining additional variance.

**Comment:**
*Line 272: The text “contributing tributaries” seems redundant. Maybe remove “contributing”. *

@@

**Response:** Fixed.

**Comment:**
*Line 395: Replace “suggest” with “such as”.*

**Response:** Fixed.

**Comment:**
*Line 420: Replace “provide” with “prove” or “be”.*

### 2023-08-11 Schramm-2023-06-PeerJ/Schramm-2023-08-PeerJ.Rmd
resubmission ready

---
title: Linking watershed nutrient loading to estuary water quality with Generalized Additive Models
preprint: false
author: 
  - name: Michael Schramm
    affiliation: 1
    corresponding: true
    email: michael.schramm@ag.tamu.edu
affiliation:
  - code: 1
    address: Texas Water Resources Institute, Texas A\&M AgriLife Research, College Station, Texas, United States
abstract: >
  Evaluating estuary water quality responses to reductions (or increases) in nutrient loading attributed to on the ground management actions can be challenging due to the strong influence of environmental drivers on nutrient loads and non-linear relationships. This study applied semiparametric approaches to calculate watershed nutrient loads and assess responses in estuary water quality to seasonally-adjusted freshwater inflow and flow-adjusted nutrient loads in Lavaca Bay, Texas. Lavaca Bay is a secondary embayment on the Texas coast displaying early potential for eutrophication and water quality degradation. Use of flow-adjusted nutrient loads allowed the study to evaluate the response in water quality to changes in nutrient loads driven by anthropogenic sources. Cross-validation indicated that, despite data constraints, semiparametric models performed well at nutrient load prediction. Based on these models, delivered annual nutrient loads varied substantially from year to year. In contrast, minimal changes in flow-normalized loads indicate that nutrient loadings were driven by natural variation in precipitation and runoff as opposed to changes in management of nonpoint sources. Models indicated no evidence of long-term changes in dissolved oxygen or chlorophyll-*a* within Lavaca Bay. However, site specific long-term increases in both organic and inorganic nitrogen are concerning for their potential to fuel eutrophication. Further analysis found freshwater inflow had strong influences on nutrient and chlorophyll-*a* concentrations but there was no evidence that changes in watershed nutrient loading explained additional variation in dissolved oxygen or chlorophyll-*a* concentrations. In addition to providing a baseline assessment of watershed nutrient loading and water quality responses in the Lavaca Bay watershed, this study provides methodological support for the use of semiparametric models in load regression models and estuary assessments.
bibliography: bibliography.bib
output:
  rticles::peerj_article:
    citation_package: biblatex
biblio-style: apa
header-includes:
    - \usepackage{csquotes}
    - \usepackage{booktabs}
    - \usepackage{longtable}
    - \usepackage{array}
    - \usepackage{multirow}
    - \usepackage{wrapfig}
    - \usepackage{float}
    - \usepackage{colortbl}
    - \usepackage{pdflscape}
    - \usepackage{tabu}
    - \usepackage{threeparttable}
    - \usepackage{threeparttablex}
    - \usepackage[normalem]{ulem}
    - \usepackage{makecell}

@@

The goal of this project is to assess watershed nutrient loading and the 
resulting water quality responses in Lavaca Bay, Texas. Lavaca Bay is a secondary bay in the larger
Matagorda Bay system located roughly halfway between Houston, Texas and Corpus
Christi, Texas. Lavaca Bay faces substantial challenges associated with
legacy contamination but general water quality parameters such as dissolved oxygen 
(DO), nutrients, and biological parameters have been well within state
water quality standards. Despite largely meeting state designated water quality 
thresholds, there have been concerning declines in abundance, biomass, and 
diversity of benthic fauna in Lavaca Bay [@beserespollackLongtermTrendsResponse2011]. 
These declines are partially attributed to reductions in freshwater inflow and 
changes in estuary salinity and are indicative of an already stressed system 
[@beserespollackLongtermTrendsResponse2011; @palmerImpactsDroughtsLow2015;
@montagnaAssessmentRelationshipFreshwater2020]. More recently, significant
linear increases in total phosphorus (TP), orthophosphate, 
total Kjeldahl nitrogen (TKN), and chlorophyll-*a* concentrations were identified
at monitoring sites within Lavaca Bay [@bugica_water_2020]. These trends in 
nutrient concentrations are concerning due to the role of nitrogen as a 
limiting factor for primary production in many Texas estuaries 
[@gardnerNitrogenFixationDissimilatory2006; @houTransformationFateNitrate2012;
@doradoUnderstandingInteractionsFreshwater2015; 
@paudelRelationshipSuspendedSolids2019; @wetz_exceptionally_2017] and the 
ramifications that changes in nitrogen loadings could have for 
productivity and eutrophication in Lavaca Bay.

@@

Water quality data for the two freshwater and three estuary locations were 
obtained from the Texas Commission on Environmental Quality (TCEQ) Surface Water 
Quality Monitoring Information System. Data submitted through the system 
are required to be collected under Quality Assurance Project Plans and lab 
method procedures outlined by the TCEQ's procedures manual. These operating 
procedures ensure consistent collection and laboratory methods are applied 
between samples collected by different entities and under different projects. 
All sites had varying time periods and availability of water quality data.
For freshwater locations, TP from January 2000 through December 2020 and 
nitrate-nitrogen (NO~3~) data from January 2005 through December 2020 were 
downloaded (Table\ \ref{tab:table1}). Less than 5-years of total
nitrogen and TKN concentration data were available at the freshwater sites and
deemed insufficient to develop load estimation models 
[@horowitzEvaluationSedimentRating2003; @snelderEstimationCatchmentNutrient2017]. 
The three estuary sites included an upper-bay site near the outlet of the
Lavaca River system (TCEQ-13563), a mid-bay site (TCEQ-13383), and the 
lower-bay site near the mouth of the Bay (TCEQ-13384). 
For estuary locations, we obtained data for TP, Nitrite+Nitrate (NO~*X*~), TKN, 
chlorophyll-*a*, and DO concentrations from January 2005 through December 2020 
(Table\ \ref{tab:table2}). All water quality data was collected approximately 
quarterly.

@@

Left-censored data were not uncommon in this dataset. Several methods are
available to account for censored data. We transformed left-censored nutrient 
concentrations to one-half the detection limit. Although this simple approach 
can introduce bias [@hornungEstimationAverageConcentration1990],  we considered 
it acceptable because high concentrations and loadings are associated with 
high-flow events and low-flow/low-concentration events will account for a small 
proportion of total loadings [@mcdowell_implications_2021].

@@

```{r fig3, error=FALSE, warning=FALSE, message=FALSE, fig.width=6.85, fig.asp=0.7, fig.cap="Aggregated estimated annual and flow-normalized annual NO\\textsubscript{3} and TP loads for USGS-08164000 and USGS-08164525.", dev='cairo_pdf'}

@@

There was no evidence of long-term changes in TP or DO concentrations at any
Lavaca Bay site (Fig.\ \ref{fig:fig5}). The upper-bay site, TCEQ-13563, had
evidence of a long-term linear increase in NO~*X*~ while chlorophyll-*a*
decreased from 2005 through 2014 (Fig.\ \ref{fig:fig5}). NO~*X*~ concentration
at the mid-bay site, TCEQ-13383, displayed a periodic pattern that is
indicative of a strong influence from inflow or precipitation. The temporal GAMs
did not provide evidence of long-term trends in any of the water quality constituents
at the lower-bay TCEQ-13384 site.

@@

Freshwater inflow provided additional explanation for changes in TP and 
NO~*X*~ concentration at all of the Lavaca Bay sites according to AIC~c~ and 
model probability values (Table\ \ref{tab:table4}).
TCEQ-13563, the site closest to the river outlet, was the only site that
had improvements in the explanations of DO and TKN concentration with the inclusion 
of inflow. Both TCEQ-13563 and TCEQ-13383, the mid-bay site, saw improvements 
in explanations for variations in chlorophyll-*a* with the inclusion of freshwater
inflow. The addition of nutrient loads (both TP and NO~3~) terms did not provide
additional explanation for changes in chlorophyll-*a* or DO concentrations.
Inclusion of TP loads provided additional explanation of TP concentrations at
the upper- and mid-bay sites, TCEQ-13563 and TCEQ-13383. Inclusion of NO~3~
loads only provided marginal improvements in the explanation of NO~*X*~ 
concentration at the lower-bay TCEQ-13384 site.

@@

Increases in freshwater inflow resulted in nearly linear increases 
in TP and NO~*X*~ concentration at all three sites (Fig.\ \ref{fig:fig6}). 
At the upper-bay TCEQ-13563 site, increases in freshwater inflow initially 
increased chlorophyll-*a* and DO concentration but concentrations leveled and
potentially decreased at higher flows. The mid-bay TCEQ-13383 site had
nearly linear increases in chlorophyll-*a* concentration in response to increased
freshwater inflow. There was no evidence of freshwater inflow effects on 
chlorophyll-*a*, TKN, or DO at the lower-bay TCEQ-13384 site.

@@

`TCEQ-13383` <- tar_read(tp_lavaca_13383_flow, store = store)

comp <- compare_smooths(model = `TCEQ-13383`,
                      `TCEQ-13384` = tar_read(tp_lavaca_13384_flow, store = store),
                      `TCEQ-13563` = tar_read(tp_lavaca_13563_flow, store = store),
                      smooths = c("s(flw_res)"),
                      overall_uncertainty = FALSE)
p_tp <- draw_smoothed_surface(comp,
                      x = flw_res,
                      xlab = "Residuals log(Inflow) [cfs]",
                      ylab =  "TP Smooth<br>Estimate [mg/L]",
                      subtitle = "A")

`TCEQ-13383` <- tar_read(no3_lavaca_13383_flow, store = store)

comp <- compare_smooths(model = `TCEQ-13383`,
                      `TCEQ-13384` = tar_read(no3_lavaca_13384_flow, store = store),
                      `TCEQ-13563` = tar_read(no3_lavaca_13563_flow, store = store),
                      smooths = c("s(flw_res)"),
                      overall_uncertainty = FALSE)
p_no3 <- draw_smoothed_surface(comp,
                      x = flw_res,
                      xlab = "Residuals log(Inflow) [cfs]",
                      ylab = "NO<sub>*x*</sub> Smooth<br>Estimate [mg/L]",
                      subtitle = "B")

>>>>

---
title: Linking watershed nutrient loading to estuary water quality with Generalized Additive Models
preprint: false
author: 
  - name: Michael Schramm
    affiliation: 1
    corresponding: true
    email: michael.schramm@ag.tamu.edu
affiliation:
  - code: 1
    address: Texas Water Resources Institute, Texas A\&M AgriLife Research, College Station, Texas, United States
abstract: >
  Evaluating estuary water quality responses to reductions (or increases) in nutrient loading attributed to on the ground management actions can be challenging due to the strong influence of environmental drivers on nutrient loads and non-linear relationships. This study applied generalized additive models to calculate watershed nutrient loads and assess responses in estuary water quality to seasonally-adjusted freshwater inflow and flow-adjusted nutrient loads in Lavaca Bay, Texas. Lavaca Bay is a secondary embayment on the Texas coast displaying early potential for eutrophication and water quality degradation. Use of flow-adjusted nutrient loads allowed the study to evaluate the response in water quality to changes in nutrient loads driven by anthropogenic sources. Cross-validation indicated that, despite data constraints, semiparametric models performed well at nutrient load prediction. Based on these models, delivered annual nutrient loads varied substantially from year to year. In contrast, minimal changes in flow-normalized loads indicate that nutrient loadings were driven by natural variation in precipitation and runoff as opposed to changes in management of nonpoint sources. Models indicated no evidence of long-term changes in dissolved oxygen or chlorophyll-*a* within Lavaca Bay. However, site specific long-term increases in both organic and inorganic nitrogen are concerning for their potential to fuel eutrophication. Further analysis found freshwater inflow had strong influences on nutrient and chlorophyll-*a* concentrations but there was no evidence that changes in watershed nutrient loading explained additional variation in dissolved oxygen and limited evidence that watershed nutrient loadings explained chlorophyll-*a* concentrations. In addition to providing a baseline assessment of watershed nutrient loading and water quality responses in the Lavaca Bay watershed, this study provides methodological support for the use of semiparametric models in load regression models and estuary assessments.
bibliography: bibliography.bib
output:
  rticles::peerj_article:
    citation_package: biblatex
biblio-style: apa
header-includes:
    - \usepackage{csquotes}
    - \usepackage{booktabs}
    - \usepackage{longtable}
    - \usepackage{array}
    - \usepackage{multirow}
    - \usepackage{wrapfig}
    - \usepackage{float}
    - \usepackage{colortbl}
    - \usepackage{pdflscape}
    - \usepackage{tabu}
    - \usepackage{threeparttable}
    - \usepackage{threeparttablex}
    - \usepackage[normalem]{ulem}
    - \usepackage{makecell}

@@

The goal of this project is to assess watershed nutrient loading and the 
resulting water quality responses in Lavaca Bay, Texas. Lavaca Bay is a secondary bay in the larger
Matagorda Bay system located roughly halfway between Houston, Texas and Corpus
Christi, Texas. Lavaca Bay faces substantial challenges associated with
legacy contamination but general water quality parameters such as dissolved oxygen 
(DO), nutrients, and biological parameters have been well within state
water quality standards. Despite largely meeting state designated water quality 
thresholds, there have been concerning declines in abundance, biomass, and 
diversity of benthic fauna in Lavaca Bay [@beserespollackLongtermTrendsResponse2011]. 
These declines are partially attributed to reductions in freshwater inflow and 
changes in estuary salinity and are indicative of an already stressed system 
[@beserespollackLongtermTrendsResponse2011; @palmerImpactsDroughtsLow2015;
@montagnaAssessmentRelationshipFreshwater2020]. More recently, significant
linear increases in total phosphorus (TP), orthophosphate, 
total Kjeldahl nitrogen (TKN), and chlorophyll-*a* concentrations were identified
at monitoring sites within Lavaca Bay [@bugica_water_2020]. While significant
nitrogen increases were only identified in TKN, both organic and
inorganic forms of nitrogen can serve as potential indicators of eutrophication
status and estuary health [@jessenMarineEutrophication2015]. Nitrogen is 
generally considered the 
limiting factor for primary production in many Texas estuaries 
[@gardnerNitrogenFixationDissimilatory2006; @houTransformationFateNitrate2012;
@doradoUnderstandingInteractionsFreshwater2015; 
@paudelRelationshipSuspendedSolids2019; @wetz_exceptionally_2017]. However, 
estuaries can display nitrogen and phosphorus co-limitation that may vary
temporally and spatially [@elserGlobalAnalysisNitrogen2007;
@conleyControllingEutrophicationNitrogen2009], 
emphasizing a need to assess a variety of nutrient species.

@@

Water quality data for the two freshwater and three estuary locations were 
obtained from the Texas Commission on Environmental Quality (TCEQ) Surface Water 
Quality Monitoring Information System. Data submitted through the system 
are required to be collected under Quality Assurance Project Plans and lab 
method procedures outlined by the TCEQ's procedures manual. These operating 
procedures ensure consistent collection and laboratory methods are applied 
between samples collected by different entities and under different projects. 
All sites had varying time periods and availability of water quality data.
For freshwater locations, TP from January 2000 through December 2020 and 
nitrate-nitrogen (NO~3~) data from January 2005 through December 2020 were 
downloaded (Table\ \ref{tab:table1}). Less than 5-years of total
nitrogen and TKN concentration data were available at the freshwater sites and
deemed insufficient to develop load estimation models 
[@horowitzEvaluationSedimentRating2003; @snelderEstimationCatchmentNutrient2017]. 
The three estuary sites included an upper-bay site near the outlet of the
Lavaca River system (TCEQ-13563), a mid-bay site (TCEQ-13383), and the 
lower-bay site near the mouth of the Bay (TCEQ-13384). 
For estuary locations, we obtained data for TP, Nitrite+Nitrate (NO~*X*~), TKN, 
chlorophyll-*a*, and DO concentrations from January 2005 through December 2020 
(Table\ \ref{tab:table2}). Freshwater nutrient samples were collect 0.3 meters 
below the surface or halfway between the surface and bottom if depth was 
insufficient. Estuary nutrient samples were all collected 0.3 meters below the 
surface. Estuary DO measurements were made at 3 locations along the water column 
(0.3 meters, mid-depth, and bottom) and averaged. All water quality data was 
collected approximately quarterly.

@@

Left-censored data were not uncommon in this dataset (4% of TP samples and 20%
of NO~3~ samples). Several methods are
available to account for censored data. We transformed left-censored nutrient 
concentrations to one-half the detection limit. Although this simple approach 
can introduce bias [@hornungEstimationAverageConcentration1990],  we considered 
it acceptable because high concentrations and loadings are associated with 
high-flow events and low-flow/low-concentration events will account for a small 
proportion of total loadings [@mcdowell_implications_2021].

@@

```{r fig3, error=FALSE, warning=FALSE, message=FALSE, fig.width=6.85, fig.asp=0.7, fig.cap="Aggregated estimated annual and flow-normalized annual NO\\textsubscript{3} and TP loads for the Lavaca (USGS-08164000) and and Navidad (USGS-08164525) Rivers.", dev='cairo_pdf'}

@@

There was no evidence of long-term changes in TP or DO concentrations at any
Lavaca Bay site (Fig.\ \ref{fig:fig5}). The upper-bay site (TCEQ-13563) had
evidence of a long-term linear increase in NO~*X*~ while chlorophyll-*a*
decreased from 2005 through 2014 (Fig.\ \ref{fig:fig5}). NO~*X*~ concentration
at the mid-bay site (TCEQ-13383) displayed a periodic pattern that is
indicative of a strong influence from inflow or precipitation. The temporal GAMs
did not provide evidence of long-term trends in any of the water quality constituents
at the lower-bay site (TCEQ-13384).

@@

Freshwater inflow provided additional explanation for changes in NO~x~ and 
chlorophyll-*a* at all of the Lavaca Bay sites according to AIC~c~ and model 
probability values (Table\ \ref{tab:table4}). Freshwater inflow also explained 
additional variation in TP at the upper- and mid-bay sites but not the lower-bay 
site. Freshwater inflow did not explain additional variation in TKN concentrations
at any of the sites. The upper- and lower-bay sites saw improvements in explanation
of DO concentration with the inclusion of inflow.

Inclusion of TP loads provided additional explanation of TP concentrations at
the upper- and mid-bay sites. Inclusion of NO~3~ loads did not provide model
explanatory improvements at any of the sites. The addition of nutrient loads 
(both TP and NO~3~) terms did not provide additional explanation for changes in 
DO concentrations but did provide model improvement for the chlorophyll-*a* model
at the mid-bay site.

Increases in aggregated freshwater inflow resulted in increases 
in TP at the upper- and mid-bay sites and increases in NO~*X*~ and 
chlorophyll-*a*concentration at all three sites (Fig.\ \ref{fig:fig6}). Linear
increases in DO concentration were observed with increasing flow at the upper-
and lower-bay sites. TKN showed no response to changes in inflow at any of the sites.

Increased TP loads resulted in nearly linear increases of TP concentration at
the upper- and mid-bay sites (Fig.\ \ref{fig:fig7}).
The relative effect size appeared much smaller than the effect of freshwater
inflow alone. Increased NO~3~ loads did effect NO~x~ concentrations at any site.
Linear increases in chlorophyll-*a* were observed in response to increased TP
loads at the mid-bay site, but not the other sites or in response to NO~3~ loads.

@@

@@

`Mid-Bay` <- tar_read(tp_lavaca_13383_flow, store = store)

comp <- compare_smooths(model = `Mid-Bay`,
                      `Lower-Bay` = tar_read(tp_lavaca_13384_flow, store = store),
                      `Upper-Bay` = tar_read(tp_lavaca_13563_flow, store = store),
                      smooths = c("s(flw_res)"),
                      overall_uncertainty = FALSE)
p_tp <- draw_smoothed_surface(comp,
                      x = flw_res,
                      xlab = "Residuals log(Inflow) [cfs]",
                      ylab =  "TP Smooth<br>Estimate [mg/L]",
                      subtitle = "A")

`Mid-Bay` <- tar_read(no3_lavaca_13383_flow, store = store)

comp <- compare_smooths(model = `Mid-Bay`,
                      `Lower-Bay` = tar_read(no3_lavaca_13384_flow, store = store),
                      `Upper-Bay` = tar_read(no3_lavaca_13563_flow, store = store),
                      smooths = c("s(flw_res)"),
                      overall_uncertainty = FALSE)
p_no3 <- draw_smoothed_surface(comp,
                      x = flw_res,
                      xlab = "Residuals log(Inflow) [cfs]",
                      ylab = "NO<sub>*x*</sub> Smooth<br>Estimate [mg/L]",
                      subtitle = "B")

`Mid-Bay` <- tar_read(chla_lavaca_13383_flow, store = store)
